Making Sequential Programs Deterministic

### 2019-04-04 design.tex
Finished with partial pass through section 2

\subsection{Maps}\label{maps}
A map is a collection of keys--value pairs.
Like other collections, a map can be nondeterministic,
order-nondeterministic, or deterministic.
Analogously to set membership, lookups are deterministic in
order-nondeterministic and deterministic maps; in addition, the iteration
order is fixed in a deterministic map.

We explain why each of these is possible by considering three common
implementation strategies.

A hash table is order-nondeterministic, because hash codes are
nondeterministic in general.

A hash table can be augmented by a linked list recording insertion order,
so that iteration order is deterministic.

A map can be backed by a sorted collection, such as a tree.  Such a map is
either \<NonDet> (if \<NonDet> keys or values are inserted into it) or
\<Det>, but never \<OrderNonDet>.

\subsection{Determinism checker specific rules for precision}\label{precision}
We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet("down") boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet
  Set}s.\todo{Explain why.}
 We added the following rule shown in \cref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det>}{\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List> }$

@@

    \caption{Improving precision}
\todo{In the \rulename{map precision} rule, should the \<Det> in the
  conclusion be $\beta_v$?}
    \label{fig-precision-rules}
\end{figure}
According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List>} and the argument to the \codeid{equals()} method is
also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have been instantiated as \<NonDet> which
is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
receiver or the argument happens to be an \codeid{OrderNonDet List}.
\todo{The following is the key!  Put in earlier.}(An order nondeterministic list does not provide any guarantee on the
equality as it is dependent on the iteration order).

Similarly, the \codeid{get()} method in the \<Map> class is annotated with polymorphic qualifiers and the rule \rulename{map precision}\ in \cref{fig-precision-rules} improves precision for \<OrderNonDet> receivers.

\subsection{Extending \ourTypeSystem}\label{extending-det-checker}
We have primarily focused on collections and annotated the corresponding classes in the JDK.
These qualifiers provide machine verifiable specifications of determinism properties.
Another source of nondeterminism that we haven't analyzed closely is the File system methods.
A program analysis developer interested in verifying determinism guarantees with respect to 
the File system can easily do so by annotating the parameter and return types of those methods in the JDK.
Alternatively, a user can write these qualifiers in a \textit{stub} file and supply it to the type checker. A \textit{stub}
only needs to contain method declarations along with their qualifiers.

>>>>

\subsection{Maps}\label{maps}
A map is a collection of key--value pairs.
Like other collections, a map or set can be nondeterministic,
order-nondeterministic, or deterministic.

It might seem that the notion of an order-nondeterministic set or map is
nonsensical, since the set and map specifications make no promises about
iteration order.  However, some subtypes do.
We explain why each of these is possible by considering three common
implementation strategies.

\begin{itemize}
\item
A hash table is order-nondeterministic, because hash codes are
nondeterministic in general.
\item
A hash table can be augmented by a linked list recording insertion order,
so that iteration order is deterministic.
\item
A map or set can be backed by a sorted collection, such as a tree.  Such a map is
either \<NonDet> (if \<NonDet> keys or values are inserted into it) or
\<Det>, but never \<OrderNonDet>.
\end{itemize}

Analogously to set membership, lookups are deterministic in
order-nondeterministic and deterministic maps; in addition, the iteration
order is fixed in a deterministic map.

\subsection{Determinism checker specific rules for precision}\label{precision}

\todo{Sets vs lists.  Set equality is independent of order.}
An order nondeterministic list does not provide any guarantee on the
equality as it is dependent on the iteration order

We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet("down") boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet
  Set}s.\todo{Explain why.}
 We added the following rule shown in \cref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det>}{\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List> }$

@@

    \caption{Special rules for set equality and map lookup.}
\todo{In the \rulename{map precision} rule, should the \<Det> in the
  conclusion be $\beta_v$?}
    \label{fig-precision-rules}
\end{figure}
According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List>} and the argument to the \codeid{equals()} method is
also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have been instantiated as \<NonDet> which
is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
receiver or the argument happens to be an \codeid{OrderNonDet List}.

Similarly, the rule \rulename{map precision}\ in \cref{fig-precision-rules}
improves precision for \<OrderNonDet> receivers.
